\vspace{-1in} Demystifying the Computer Science \\PhD Admission in US Universities\\{\large A guide for International Students

### 2023-08-13 phd-cs-us.tex
apply comments from Hung Le.

\newcommand{\mycomment}[3][\color{blue}]{{#1{{#2}: {#3}}}}
\newcommand{\tvn}[1]{\mycomment{TVN}{#1}}{}
\newcommand{\didi}[1]{\mycomment{Didier}{#1}}{}
\newcommand{\tl}[1]{\mycomment{ThanhLe}{#1}}{}
\newcommand{\red}[1]{{\color{red}{#1}}}
\renewcommand{\figurename}{Fig.}
\renewcommand{\tablename}{Tab.}
\title{\vspace{-1in} Demystifying the Computer Science \\PhD Admission in US Universities\\{\large A guide for Vietnamese and International Students}}
%\date{\today}
\author{\small \href{https://nguyenthanhvuh.github.io}{ThanhVu (Vu) Nguyen} (\href{mailto:tvn@gmu.edu}{tvn@gmu.edu})}

@@

\begin{abstract}
Having been involved in PhD admissions for many years, I've
realized that many \textbf{international students}, especially those from smaller countries such as \emph{Vietnam}, lack a clear understanding of
the Computer Science PhD admission process at US universities. This confusion not only
discourages students from applying but also creates the perception that
getting admitted is difficult compared to CS PhD programs in other countries.

So I want to share some details about the PhD admission process and advice for those who are interested in applying for a \textbf{PhD in Computer Science in the US}.
While this document is primarily intended for Vietnamese interested in computer science, it may be relevant to students from various countries and disciplines.
Moreover, although many examples are specifics for George Mason University, the information should be generalizable to other R1\footnote{An \href{https://en.wikipedia.org/wiki/List_of_research_universities_in_the_United_States}{R1 institution} in the US is a research-intensive university with a high level of research activity across various disciplines.} institutions in the US  (though \emph{very} top schools could be very selective, e.g., see the \href{https://da-data.blogspot.com/2015/03/reflecting-on-cs-graduate-admissions.html}{admission process} at CMU).

@@

Then applications will be reviewed by a
\textbf{PhD admission committee} that consists of faculty members in CS (in some cases the committee can involve affiliated faculty from different disciplines). These faculty have a wide-range of expertise and background to ensure diverse perspectives in the evaluation process. The size and the review load of the committee depends on the department size. At GMU, the PhD committee typically has 15--20 faculty, and each committee member is assigned from 25--30 applications. Note that at GMU (and probably many other schools) we have a separate committee for the MS program.

@@

@@

@@

\begin{wrapfigure}{r}{0.47\linewidth}
      \begin{center}
        \includegraphics[width=1\linewidth]{c6.png}
      \end{center}
\end{wrapfigure}
Many students have letters written by themselves and signed by their professors. These have little
values. We can recognize them and will consider them weakness.
Similarly, many professors write generic letters for students, e.g., the students didn't do any
research or make any impression for the professor to write about. These letters are also not useful and considered weak.

@@

@@

\begin{wrapfigure}{r}{0.52\linewidth}
    \begin{center}
      \includegraphics[width=1\linewidth]{c2.png}
    \end{center}
  \end{wrapfigure}
While you might not have control over LORs \hyperref[sec:your-school]{where your go to school}, you do over your
statement! So write it well because we take it seriously.
In your statement, you have the opportunity to make  make your application stand out and unique, which can make you \emph{fit} the CS program you're applying to, even if you don't have very strong research experience.
A well-written LOR also shows that you can communicate, which is very important in research, and that you can effectively teach and communicate with students, which is important for TA funding (see \S\ref{sec:funding}).
%is important because if you need (GTA) funding, it will provide evidence
%that you can teach and communicate with students.

There are various guides on writing statement, e.g.,~\cite{blattman2022writing}, and \href{https://cs-sop.org/}{many example statements} are available. So I will not talk too much about statements. In short, discuss about your research vision and convince us that you can achieve it through your experience, e.g., published papers, or if you work on some projects by yourself, talk about it. Also, use the statement to talk about stuff that admission committee members might not know about, e.g., your Github project with 1K+ stars or  your regular contributions to well-known open-source projects.

@@

@@

This can be an issue for students in many top international universities where the competition is so high that very good students students can still have low ranking (and be overlooked by Admission committee).
So what to do with this? well, same as \hyperref[sec:your-school]{before}, e.g., put a note about this in your statement and ask your LoR writers to talk about it.

@@

\subsection{Interview}

Sometime a faculty wants to interview an applicant to make a decision. This means they are leaning toward admitting you (if we don't like your application, we will not bother doing the interview).

An interview lasts about 15--30 mins, and one implicit thing you will be evaluated on is whether you can communicate effectively (i.e., speak/understand English).  You will also get chance to ask questions about the university so think of something to ask (just the same as interviewing at a company).

\begin{commentbox}[Vu]
At GMU, we are encouraged to interview candidates. For very strong candidates, the interview is actually to recruit them.  In some cases a faculty interviews a candidate that they see potentials and wants to argue for admission, i.e., without the interview, that application is likely rejected. In any case, getting interview means you have a very good chance of being admitted.
\end{commentbox}

@@

Around March you should hear back from most PhD programs that you applied to (if not, send email and ask).
If you have offers, congratulations!  Now you're at a different game because the schools that admit you will now try to get you to accept them!   You will likely have to make your decision by around April 15.

@@

 Show something that makes you \textbf{stand out}, e.g., are you a female or a minority in CS (research for this on Google)? Do you participate in outreach activities that help increase diversity and inclusion in CS?  All of these are important in CS in the US.

@@

\begin{wrapfigure}{r}{0.48\textwidth}
      \begin{center}
        \includegraphics[width=0.48\textwidth]{c8.png}
      \end{center}
    \end{wrapfigure}
The short answer is that tenured-track faculty (e.g., assistant professors) are more likely to be young and active in research (they have to, in order to get tenure). Thus, they will likely have more time to work with you and push you to do research and publish. However, they may not have as much experience in managing students and may not have as much funding as tenured faculty.

Tenured faculty (e.g., associate and full profs.) are more likely to be older, more well-known, and have more experience in managing students.  However, they might not push you as hard (they don't have to, they already got tenured) and expect you to figure things out yourself (so you need to be very independent).  Some tenured faculty are also no longer active in research and more involved with administrative duties or with their startup companies.

@@

\subsection{Should you contact a US professor? How to do so ``correctly''?}

Faculty received many "cold" e-mails from international students seeking for admission, TA, and RA. Most of the time, we ignore these emails; but in rare occasion we do answer them. So how to write an email that can get our attention?

@@

I read your TSE'21 paper on numerical invariant generation, and I am interested in this line of dynamic invariant research. I have worked (optional: with prof. Y at Z) on static program analysis and I think it could be used to tackle the spurious issues mentioned in your paper. I have a small paper at conference/workshop C and a project on symbolic execution at Github G.

@@

\subsection{How long does it take to complete the PhD program?}

\begin{wrapfigure}{r}{0.28\textwidth}
      \begin{center}
        \includegraphics[width=0.28\textwidth]{c4a.png}
      \end{center}
    \end{wrapfigure}
    Typically, 5--7 years for PhD in CS in the US.  This can be longer than CS PhD at universities in other countries, which might require MS first (recall that US PhD programs do not require MS and you can get MS along the way to PhD). Within these 5--7 years, CS PhD students in the US also often take a ``leave of absence'' for 1--2 semesters to do internship at companies and research labs.

@@

@@

\begin{wrapfigure}{l}{0.5\textwidth}
      \begin{center}
        \includegraphics[width=0.5\textwidth]{c5.png}
      \end{center}
    \end{wrapfigure}
If you don't know the professor (e.g., first email contact), then use \textbf{Prof. Lastname} or \textbf{Dr. Lastname}. I've seen many international students write \textbf{Prof.} or \textbf{Dr.} \textbf{Firstname Lastname}.  Writing like that makes it like you copy and paste the names, so no need to do so,  just Prof. or Dr. Lastname.

@@

I started writing this document in May 2023 and have been updating it since then (mostly around deadline time when I tend to procastinate :p). I have shared it with many students and faculty members and received many valuable feedback.  I also hope that it will be updated and improved by many people in the future.  I have put the source code of this document on \href{https://github.com/nguyenthanhvuh/phd-cs-us}{GitHub} so that anyone can contribute to it.  I also hope that this document will be translated to Vietnamese and other languages so that it can reach more students.

\paragraph{Acknowledgement} Many people have contributed to this document.
Craig Yu (GMU) and Hakan Aydin (GMU) provided valuable input in the early version. Other GMU faculty members also have provided feedback and contributions.  Many students including Didier (GMU), Thanh (Melbourne), Dat (Melbourne) have contributed valuable questions and feedback. 
\textbf{Thank you!}

>>>>

\newcommand{\mycomment}[3][\color{blue}]{{#1{{#2}: {#3}}}}
\newcommand{\tvn}[1]{\mycomment{TVN}{#1}}{}
\newcommand{\didi}[1]{\mycomment{Didier}{#1}}{}
\newcommand{\tl}[1]{\mycomment{ThanhLe}{#1}}{}
\newcommand{\red}[1]{{\color{red}{#1}}}
\renewcommand{\figurename}{Fig.}
\renewcommand{\tablename}{Tab.}
\title{\vspace{-1in} Demystifying the Computer Science \\PhD Admission in US Universities\\{\large A guide for International Students}}
%\date{\today}
\author{\small \href{https://nguyenthanhvuh.github.io}{ThanhVu (Vu) Nguyen} (\href{mailto:tvn@gmu.edu}{tvn@gmu.edu})}

@@

\begin{abstract}
Having been involved in PhD admissions for many years, I've
realized that many \textbf{international students}, especially those from smaller countries such as Vietnam, lack a clear understanding of
the Computer Science PhD admission process at US universities. This confusion not only
discourages students from applying but also creates the perception that
getting admitted is difficult compared to CS PhD programs in other countries.

So I want to share some details about the PhD admission process and advice for those who are interested in applying for a \textbf{PhD in Computer Science in the US}.
While this document is primarily intended for students in Vietnam interested in CS, it may be relevant to students from various countries and disciplines.
Moreover, although many examples are specifics for George Mason University, the information should be generalizable to other R1\footnote{An \href{https://en.wikipedia.org/wiki/List_of_research_universities_in_the_United_States}{R1 institution} in the US is a research-intensive university with a high level of research activity across various disciplines.} institutions in the US  (though \emph{very} top schools could be very selective, e.g., see the \href{https://da-data.blogspot.com/2015/03/reflecting-on-cs-graduate-admissions.html}{admission process} at CMU).

@@

Then applications will be reviewed by a
\textbf{PhD admission committee} that consists of faculty members in CS (in some cases the committee can involve affiliated faculty from different disciplines). These faculty have a wide-range of expertise and background to ensure diverse perspectives in the evaluation process. The size and the review load of the committee depends on the department size. At GMU, the PhD committee typically has 15--20 faculty, and each committee member is assigned from 25--30 applications. Note GMU and UMass (and probably many other schools) have a separate committee for the MS program.

@@

\begin{commentbox}[Hung]
    The application fee varies between schools, and is about \$70--\$100. I found this number pretty ridiculous. In most cases, there is no proper justification for that amount. Yes, the application fee could act as a filter, but a nominal amount would do the job. While a blanket waiver might not be feasible, universities should extend the eligibility of fee waiver applications to more countries.
\end{commentbox}

@@

\begin{commentbox}[Hung]
The reputation of VinAI has been increasing steadily over the past few years; many of my colleagues heard about VinAI.
\end{commentbox}

@@

\begin{wrapfigure}{r}{0.47\linewidth}
\vspace{-0.2in}
      \begin{center}
        \includegraphics[width=1\linewidth]{c6.png}
      \end{center}
      \vspace{-0.1in}
\end{wrapfigure}
Many students have letters written by themselves and signed by their professors. These have little
values. We can recognize them and will consider them weakness.
Similarly, many professors write generic letters for students, e.g., the students didn't do any
research or make any impression for the professor to write about. These letters are also not useful and considered weak.

@@

\begin{commentbox}[Hung]
A sad reality is that most professors in Vietnam \textbf{DO NOT} know how to write a good letter, or are lazy in writing letters hence delegate the writing to the students. Unfortunately, there is no easy solution to this problem.
\end{commentbox}

@@

\begin{wrapfigure}{r}{0.52\linewidth}
\vspace{-0.2in}
    \begin{center}
      \includegraphics[width=1\linewidth]{c2.png}
    \end{center}
    \vspace{-0.2in}
  \end{wrapfigure}
While you might not have control over LORs \hyperref[sec:your-school]{where your go to school}, you do over your
statement! So write it well because we take it seriously.
In your statement, you have the opportunity to make  make your application stand out and unique, which can make you \emph{fit} the CS program you're applying to, even if you don't have very strong research experience.
A well-written LOR also shows that you can communicate, which is very important in research, and that you can effectively teach and communicate with students, which is important for TA funding (see \S\ref{sec:funding}).
%is important because if you need (GTA) funding, it will provide evidence
%that you can teach and communicate with students.

There are various guides on writing statement, e.g.,~\cite{blattman2022writing}, and \href{https://cs-sop.org/}{many example statements} are available. So I will not talk too much about statements. In short, discuss about your research goal and vision, and convince us that you can achieve it through your experience, e.g., published papers, or if you work on some projects by yourself, talk about it. Also, use the statement to talk about stuff that admission committee members might not know about, e.g., your Github project with 1K+ stars or  your regular contributions to well-known open-source projects.

\begin{commentbox}[Hung]
I think asking for a “research vision” from a Ph.D. applicant is too much. Even people graduated with a Ph.D. have a great difficulty in explicating their researh vision. In my opinion, you should focus on showing (not telling) your research passion (if you do have).
\end{commentbox}

@@

\begin{commentbox}[Hung]
I personally don’t care about which schools students attended. There are many other ways to evaluate an applicant.
\end{commentbox}

@@

This can be an issue for students in many top international universities where the competition is so high that very good students students can still have low ranking from these schools (and be overlooked by Admission committee).
So what to do with this? well, same as \hyperref[sec:your-school]{before}, e.g., put a note about this in your statement and ask your LoR writers to talk about it.

@@

\subsection{Interview}
Sometimes a faculty wants to interview an applicant to make a decision. Typically, this means that they are considering admitting you. If they are not interested in your application, they will not proceed with an interview.

An interview often lasts about 15-30 minutes, and one important aspect of evaluation is your ability to effectively communicate, including speaking and understanding English. In addition, during the interview, you will have the opportunity to ask questions about the university. It's essential to prepare some thoughtful questions, similar to interviewing at a company.

\begin{commentbox}[Vu]
At GMU, we are encouraged to interview candidates. For very strong candidates, the interview is actually to recruit them.  In some cases a faculty interviews a candidate that they see potentials and want to advocate for their admission. Without the interview, such applications may be more likely to be rejected. Overall, being selected for interview means a very good chance of being admitted.
\end{commentbox}

@@

By around March you should hear back from most PhD programs you applied to . If you haven't heard back, reach our via email and ask about the status of your application.
If you receive offers, congratulations!  Now you're at a different game because the schools that have admitted you will try to get you to accept them!   You will likely have to make your decision by around April 15.

@@

 Show something that makes you \textbf{stand out}, e.g., do you have a degree or background in music and want to integrate that to CS? are you a female or a minority in CS (research for this on Google)? Do you participate in outreach activities that help increase diversity and inclusion in CS? All of these are unique and would get noticed from reviewers.

@@

\begin{wrapfigure}{r}{0.48\textwidth}
\vspace{-0.2in}
      \begin{center}
        \includegraphics[width=0.48\textwidth]{c8.png}
      \end{center}
      \vspace{-0.2in}
    \end{wrapfigure}
The short answer is that tenured-track faculty (e.g., assistant professors) are more likely to be young and active in research (they have to, in order to get tenure). Thus, they will likely have more time to work with you and push you to do research and publish. However, they may not have as much experience in managing students and may not have as much funding as tenured faculty.

Tenured faculty (e.g., associate and full profs.) are more likely to be older, more well-known, and have more experience in managing students.  However, they might not push you as hard (they don't have to, they already got tenured). They might also expect you to figure things out yourself (so you need to be very independent).  Some tenured faculty are also no longer active in research and more involved with administrative duties or with their startup companies.

@@

\subsection{Should you contact a US professor? What to do to get a desired reply?}

Faculty received many "cold" e-mails from international students seeking for admission, TA, and RA. Most of the time, we ignore these emails, but in some rare occasions we do answer them. So how to write an email that can get our attention?

@@

I am writing to inquire about potential research opportunities as a GRA in your group at GMU. Currently I am an undergraduate student in Computer Science at UNIV and plan to graduate in May 2023.

I have read your TSE'21 paper on numerical invariant generation, and I am interested in this line of dynamic invariant research. I have worked (optional: with prof. Y at Z) on static program analysis and I think it could be used to tackle the spurious issues mentioned in your paper. I have a small paper at conference/workshop C and a project on symbolic execution at Github G.

@@

\subsection{How long does it take to complete the CS PhD program?}

\begin{wrapfigure}{r}{0.28\textwidth}
\vspace{-0.2in}
\begin{center}
\includegraphics[width=0.28\textwidth]{c4a.png}
\end{center}
\vspace{-0.1in}
\end{wrapfigure}
    Typically, 5--7 years for PhD in CS in the US.  This can be longer than CS PhD at universities in other countries, which might require MS first (recall that US PhD programs do not require MS and you can get MS along the way to PhD). Within these 5--7 years, CS PhD students in the US also often take a ``leave of absence'' for 1--2 semesters to do internship at companies and research labs.

@@

\subsection{PhD in other related field such as CE, IST, Cybersecurity, and Stats}

In many cases you do not need to do a PhD in CS to work in your area of interested. For example, in addition to having a traditional CS department, GMU has IST and Cybersecurity departments with faculty who have PhD in CS and work on CS topics (e.g., AI, Security, Robotics).  In short, it is totally possible that you still get to do CS research and publish in CS-related venues even if you're not in a traditional CS program.  It is very common to see faculty with CS PhD in a non-CS department as well as faculty with non-CS PhD in CS department.

It is worth it to note that if your intention is a PhD in CS, then you likely need to be in the CS dept \emph{and} advised by a CS faculty. For example, if your advisor is not in CS, then you likely cannot not get a CS degree. In fact, if a faculty is not in CS, they likely cannot be the main advisor of a CS PhD student---they may \emph{co-advise} or be in the PhD dissertation committee, but your main advisor will need to be a tenured or tenure-track faculty in CS. If in doubt, you should check with the CS department for their requirements.

Note that for the purpose of helping students interested in a PhD in CS, CSRankings only includes tenured or tenured track faculty who can advise CS PhD students. I also have compiled a \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Viet-CS-Profs-US}{list} of Vietnamese faculty who can advise PhD students.

@@

\begin{wrapfigure}{l}{0.5\linewidth}
\vspace{-0.3in}
      \begin{center}
        \includegraphics[width=1.0\linewidth]{c5.png}
      \end{center}
      \vspace{-0.3in}
    \end{wrapfigure}
If you don't know the professor (e.g., first email contact), then use \textbf{Prof. Lastname} or \textbf{Dr. Lastname}. I've seen many international students write \textbf{Prof.} or \textbf{Dr.} \textbf{Firstname Lastname}.  Writing like that makes it like you copy and paste the names, so no need to do so,  just Prof. or Dr. Lastname.

@@

I started writing this document in May 2023 and have been updating it since then (mostly around deadline time when I tend to procrastinate!). I have put the source code of this document on \href{https://github.com/nguyenthanhvuh/phd-cs-us}{GitHub} so that anyone can contribute to it.

\paragraph{Acknowledgement} Many people have contributed to this document.
Craig Yu (GMU), Hakan Aydin (GMU), Hung Le (UMass) provided valuable input in the early version. Other GMU faculty members also have provided feedback and contributions.  Many students including Didier (GMU), Thanh (Melbourne), and Dat (Melbourne) have contributed valuable questions and feedback. 
\textbf{Thank you!}
